Towards atomistic resolution structure of phosphatidylcholine headgroup and glycerol backbone at different ambient conditions\footnote{Publication about results presented in the NMRlipids project}

### 2015-10-11 HGmodelMANUSCRIPT/HGmodel_ACStemplate.tex
Tuned language.

\subsection{Response to dehydration and cholesterol content}
In addition to pure phosphatidylcholine bilayers at full hydration, the choline headgroup order parameters
have been measured under various different conditions~\cite{gally75,brown77,brown78,akutsu81,altenbach84,scherer89,bechinger91,ulrich94,dvinskikh05b,castro08,kapla12,ferreira13}.
Also the order parameters for the glycerol backbone have been measured with $^1$H-$^{13}$C NMR in dehydrated conditions~\cite{dvinskikh05b}, and as a function 
of anesthetics~\cite{castro08} and glycolipids~\cite{kapla12} for DMPC, and as a function of cholesterol 
concentration for POPC~\cite{ferreira13}. Due to the high resolution in the NMR (especially $^2$H NMR) experiments,
even very small order parameter changes resulting from the varying conditions can be measured 
(see Ref.~\citenum{accuracyPOST} for more discussion.) 
However, as already discussed above, it is not simple to deduce 
the structural changes from order parameter changes~\cite{akutsu91,semchyschyn04}. Consequently, comparison of the order parameters
between simulations and experiments in different conditions can be used to assess the quality of the force field 
in different situations, and, if the quality is good, to interpret the structural changes in experiments.
Here we exemplify such comparison for a lipid bilayer under low hydration levels and when varying amounts of cholesterol is included in the bilayer. 
The interaction between ions and a phosphatidylcholine bilayer will be discussed in a separate study~\cite{ionpaper}.

\subsubsection{Phospholipid bilayer with low hydration level}
Fig.~\ref{ordPhydr} shows the published~\cite{dvinskikh05b,ulrich94,bechinger91} experimental order parameters
for the glycerol backbone and choline as a function of hydration level.
Despite slight differences in temperature and acyl chain composition,
the three independently reported data sets for the choline ($\beta$ and $\alpha$) segments agree well with each other:
Both order parameters increase with decreasing hydration level.
The glycerol backbone order parameters (g$_{3}$, g$_{2}$, g$_{1}$), in contrast,
have been observed~\cite{dvinskikh05b} to slightly decrease with dehydration.
Note that the original experiments~\cite{dvinskikh05b,ulrich94,bechinger91} measured only 
absolute values, but here
we included the signs measured in separate studies~\cite{hong95a,hong95b,gross97}. 
Consequently, the negative $\beta$ order parameter actually increases with dehydration 
as its absolute value decreases~\cite{dvinskikh05b,ulrich94,bechinger91}.
\begin{figure}[]
  \centering
  \includegraphics[width=7.3cm]{../DATAreportediINblog/dehydration.pdf}
    \caption{\label{ordPhydr}
    The effect of dehydration on glycerol and choline order parameters in experiments.
    The magnitudes of order parameters are measured for DMPC ($^1$H-$^{13}$C NMR) at 314~K~\cite{dvinskikh05b}, 
    for POPC ($^2$H NMR) at 296~K~\cite{bechinger91} and for DOPC ($^2$H NMR) at 303~K~\cite{ulrich94}. 
    The signs are based on the measurements by Hong et al.~\cite{hong95a,hong95b} 
    and Gross et al.~\cite{gross97}.
    Note that to elucidate the relative change as a function of hydration level, the simulation results were vertically shifted; 
    the shift magnitudes for each of the force fields are listed (SCH+shift) in the y-label.
  }
\end{figure}

@@

We note that the modest ($\Delta S_\mathrm{CH}<0.02$ for g$_1$; $<0.04$ for g$_2$, g$_3$; see Fig.~\ref{ordPchol})
effects of cholesterol
on the glycerol backbone order parameters of POPC
measured by $^1$H-$^{13}$C NMR~\cite{ferreira13}
agree well with the results for phosphatidylethanolamine (PE) measured by $^{2}$H NMR~\cite{ghosh82}.
This further supports the ideas that the glycerol backbone structural behaviour is independent of the
headgroup composition~\cite{gally81} and that the headgroup structure is largely independent of the acyl chain region content unless
charges are present~\cite{scherer87}.
\begin{figure}[]
  \centering
  \includegraphics[width=7.3cm]{../DATAreportediINblog/cholesterolization_lp.pdf}
%    \includegraphics[width=7.3cm]{../DATAreportediINblog/cholesterolization_smooth.pdf}
  \caption{\label{ordPchol}
    The effect of cholesterol content on the glycerol backbone and choline order parameters in experiments~\cite{brown78,ferreira13} and simulations
    with the Berger-POPC-07/H\"oltje-CHOL-13, CHARMM36 and MacRog force fields. The signs in the experimental values are based on the measurements by Hong et al.~\cite{hong95a,hong95b} 
    and Gross et al.~\cite{gross97}.  Most order parameters from Berger-POPC-07/H\"oltje-CHOL-13 model for g$_1$ are beoynd the y-axis scale.
    In order to elucidate the relative change as a function of cholesterol content, the simulation results were vertically shifted; 
    the shift magnitudes for each of the force fields are listed (SCH+shift) in the y-label.
}
\end{figure}

@@

(4) Only modest changes of glycerol backbone and choline order parameters are observed experimentally with increasing cholesterol content.
When interpreted using the computational lipid model that we found to have the most realistic response to cholesterol,
this observation means that cholesterol induces only minor changes in (the g$_3$-g$_2$-g$_1$-O(sn-1)
dihedral of the glycerol backbone, in other words, there is no major conformational change of the lipid.

@@

We acknowledge all the discussion participants at nmrlipids.blogspot.fi.
AB and CL acknowledges financial support from the French National Research Agency
(ANR: “Biolubrication by phospholipid membranes” Bioub2012) and
computing time allocation from P{\^o}le Scientifique de Mod{\'e}lisation Num{\'e}rique from the ENS Lyon (PSMN),
and Centre Informatique National de l'Enseignement Sup{\'e}rieur (CINES, Montpelier, France)
(Project c2015096850).
FFR acknowledges CONACYT and DGAPA UNAM IG100513
for financial support, Cluster H\'ibrido de Superc\'omputo Xiuhcoatl - CINVESTAV and Miztli - UNAM for computational resources. 
MJ and JT acknowledge and CSC - IT Center for Science for computational resources (project number tty3979).
MJ also acknowledges the Finnish Doctoral Programme in Computational Sciences (FICS) for funding.
JT, MJ, TR and WK acknowledge the funding from the Academy of Finland (Centre of Excellence program) 
and the European Research Council (Advanced Grant project CROWDED-PRO-LIPIDS).
WK acknowledges CSC -- IT Centre for Science (Espoo, Finland) for excellent computational resources (project number tty3995).
MSM acknowledges financial support from the Volkswagen Foundation (86110).
LM acknowledges funding proved by the Institut national de la sant\'e et de la recherche m\'edicale (INSERM).
JM acknowledges CSC -- IT Center for Science for computational resources.
OHSO acknowledges Tiago Ferreira and Paavo Kinnunen for useful discussions, 
the Emil Aaltonen foundation for financial support, 
Aalto Science -- IT project and CSC -- IT Center for Science for computational resources. 
HS acknowleges Catherine Etchebest and St\'ephane T\'eletch\'ea for useful discussions and continued support, the HPC 
resources granted from GENCI-CINES (Grant 2014-c2014077209) and computer facilities provided by R\'egion Ile de France and INTS (SESAME 2009 project).

>>>>

\subsection{Response to dehydration and cholesterol content}
In addition to pure phosphatidylcholine bilayers at full hydration, the choline headgroup order parameters
have been measured under various different conditions~\cite{gally75,brown77,brown78,akutsu81,altenbach84,scherer89,bechinger91,ulrich94,dvinskikh05b,castro08,kapla12,ferreira13}.
Also the order parameters for the glycerol backbone have been measured with $^1$H-$^{13}$C NMR in dehydrated conditions~\cite{dvinskikh05b}, and as a function 
of anesthetics~\cite{castro08} and glycolipids~\cite{kapla12} for DMPC, and as a function of cholesterol 
concentration for POPC~\cite{ferreira13}. Due to the high resolution in the NMR (especially $^2$H NMR) experiments,
even very small order parameter changes resulting from the varying conditions can be measured 
(see Ref.~\citenum{accuracyPOST} for more discussion)---but,
as already discussed above, it is not simple to deduce 
the structural changes from order parameter changes~\cite{akutsu91,semchyschyn04}. However, comparison of the order parameters
between simulations and experiments in different conditions can be used to assess the quality of the force field 
in different situations, and, if the quality is good, to interpret the structural changes in experiments.
Here we exemplify such comparison for a lipid bilayer under low hydration levels and when varying amounts of cholesterol is included in the bilayer. 
The interaction between ions and a phosphatidylcholine bilayer will be discussed in a separate study~\cite{ionpaper}.

\subsubsection{Phospholipid bilayer with low hydration level}
Fig.~\ref{ordPhydr} shows the published~\cite{dvinskikh05b,ulrich94,bechinger91} experimental order parameters
for the glycerol backbone and choline as a function of hydration level.
Despite slight differences in temperature and acyl chain composition,
the three independently reported data sets for the choline ($\beta$ and $\alpha$) segments agree well with each other:
Both order parameters increase with decreasing hydration level.
The glycerol backbone order parameters (g$_{3}$, g$_{2}$, g$_{1}$), in contrast,
have been observed~\cite{dvinskikh05b} to slightly decrease with dehydration.
Note that the original experiments~\cite{dvinskikh05b,ulrich94,bechinger91} measured only 
absolute values, but here
we included the signs measured in separate studies~\cite{hong95a,hong95b,gross97}. 
Consequently, the negative $\beta$ order parameter actually increases with dehydration 
as its absolute value decreases~\cite{dvinskikh05b,ulrich94,bechinger91}.
\begin{figure}[]
  \centering
  \includegraphics[width=7.3cm]{../DATAreportediINblog/dehydration.pdf}
    \caption{\label{ordPhydr}
    The effect of dehydration on glycerol and choline order parameters in experiments.
    The magnitudes of order parameters are measured for DMPC ($^1$H-$^{13}$C NMR) at 314~K~\cite{dvinskikh05b}, 
    for POPC ($^2$H NMR) at 296~K~\cite{bechinger91} and for DOPC ($^2$H NMR) at 303~K~\cite{ulrich94}. 
    The signs are based on the measurements by Hong et al.~\cite{hong95a,hong95b} 
    and Gross et al.~\cite{gross97}.
    Note that to elucidate the relative change as a function of hydration level, the simulation results were vertically shifted; 
    the shift magnitudes for each of the force fields are listed (S$_\mathrm{CH}$+shift) in the y-label.
  }
\end{figure}

@@

We note that the modest ($\Delta S_\mathrm{CH}<0.02$ for g$_1$; $<0.04$ for g$_2$, g$_3$; see Fig.~\ref{ordPchol})
effects of cholesterol
on the glycerol backbone order parameters of POPC
measured by $^1$H-$^{13}$C NMR~\cite{ferreira13}
agree well with the results for phosphatidylethanolamine (PE) measured by $^{2}$H NMR~\cite{ghosh82}.
This further supports the ideas that the glycerol backbone structural behaviour is independent of the
headgroup composition~\cite{gally81} and that the headgroup structure is largely independent of the acyl chain region content unless
charges are present~\cite{scherer87}.
\begin{figure}[]
  \centering
  \includegraphics[width=7.3cm]{../DATAreportediINblog/cholesterolization_lp.pdf}
%    \includegraphics[width=7.3cm]{../DATAreportediINblog/cholesterolization_smooth.pdf}
  \caption{\label{ordPchol}
    The effect of cholesterol content on the glycerol backbone and choline order parameters in experiments~\cite{brown78,ferreira13} and simulations
    with the Berger-POPC-07/H\"oltje-CHOL-13, CHARMM36 and MacRog force fields. The signs in the experimental values are based on the measurements by Hong et al.~\cite{hong95a,hong95b} 
    and Gross et al.~\cite{gross97}.  %Most order parameters from Berger-POPC-07/H\"oltje-CHOL-13 for g$_1$ are beyond the y-axis scale.
    In order to elucidate the relative change as a function of cholesterol content, the simulation results were vertically shifted; 
    the shift magnitudes for each of the force fields are listed (S$_\mathrm{CH}$+shift) in the y-label.
}
\end{figure}

@@

(4) Only modest changes of glycerol backbone and choline order parameters are observed experimentally with increasing cholesterol content.
When interpreted using the computational lipid model that we found to have the most realistic response to cholesterol,
this observation means that cholesterol induces only minor changes in the g$_3$-g$_2$-g$_1$-O(sn-1)
dihedral of the glycerol backbone, in other words, there is no major conformational change of the lipid.

@@

We acknowledge all the discussion participants at \url{nmrlipids.blogspot.fi}.
AB and CL acknowledge financial support from the French National Research Agency
(ANR: Biolubrication by phospholipid membranes, Biolub2012) and
computing time allocation from P{\^o}le Scientifique de Mod{\'e}lisation Num{\'e}rique from the ENS Lyon (PSMN),
and Centre Informatique National de l'Enseignement Sup{\'e}rieur (CINES, Montpelier, France)
(Project c2015096850).
FFR acknowledges CONACYT and DGAPA UNAM IG100513
for financial support, Cluster H\'ibrido de Superc\'omputo Xiuhcoatl - CINVESTAV and Miztli - UNAM for computational resources. 
MJ and JT acknowledge and CSC -- IT Center for Science for computational resources (project number tty3979).
MJ also acknowledges the Finnish Doctoral Programme in Computational Sciences (FICS) for funding.
JT, MJ, TR and WK acknowledge the funding from the Academy of Finland (Centre of Excellence program) 
and the European Research Council (Advanced Grant project CROWDED-PRO-LIPIDS).
WK acknowledges CSC -- IT Centre for Science (Espoo, Finland) for excellent computational resources (project number tty3995).
MSM acknowledges financial support from the Volkswagen Foundation (86110).
LM acknowledges funding proved by the Institut national de la sant\'e et de la recherche m\'edicale (INSERM).
JM acknowledges CSC -- IT Center for Science for computational resources.
OHSO acknowledges Tiago Ferreira and Paavo Kinnunen for useful discussions, 
the Emil Aaltonen foundation for financial support, 
Aalto Science -- IT project and CSC -- IT Center for Science for computational resources. 
HS acknowleges Catherine Etchebest and St\'ephane T\'eletch\'ea for useful discussions and continued support, the HPC 
resources granted from GENCI-CINES (Grant 2014-c2014077209) and computer facilities provided by R\'egion Ile de France and INTS (SESAME 2009 project).
